---
title: "Programming in R"
output:
  github_document:
    md_extension: +gfm_auto_identifiers
    toc: true
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r echo=FALSE, message=FALSE}
knitr::opts_chunk$set(message=FALSE, comment="")
```

## Getting Started
R is a free programming language for statistical computing and graphics. It is an implementation of the S programming language and was created by Ross Ihaka and Robert Gentleman at the University of Auckland, New Zealand. R is currently developed by the R Development Core Team. RStudio is an Integrated Development Environment (IDE) for R.

## The Basics

We can use R as a calculator to do simple math operations such as addition `+`, subtraction `-`, multiplication `*`, division `\`, and raise a value to a power `^`. We can type these calculations in the console or run them in an R script that extension ends in `.R` 

```{r}
# We can use hashtags to make comments about our code
```

```{r}
4 + 5
4 - 5
4 * 5
4 / 5
4 ^ 5
```

### Data Structures

R has many data structures and types that we can use, depending on the information we want to work with. 
The major data types include:

- character
- numeric (real or decimal)
- integer
- logical
- double
- complex

The major data structures include:

- Scalars
- Atomic Vectors
- Factors
- Lists
- Matrices and Arrays
- Dataframes

### Scalars

The simplest type of object is a scalar which is an object with one value. We can assign a value or calculations to a variable using the assignment operator `<-`. 

Note: The equals sign `=` is not an assignment operator in R and has a different functionality which will be discussed further below.

To create scalar data objects x and y:

```{r}
x <- 4    
y <- 5
```

The objects x and y were set a numeric data type.

We can manipulate these objects in R and perform different calculations together. To print the value of these variables, we can use the print() function or call the variable itself.

```{r}
z <- x+y   
z
print(z)
x*y/z
```

As stated above, we can also create data objects of other data types such as logical and character mode.

For logical data, we use TRUE (T) and FALSE (F)

```{r}
Logical <- T
Logical
```

For character data, we use single or double quotation to enclose the data

```{r}
Character_Data <- "T"
Character_Data
```

We can use available functions in R to determine the mode or type of data we are working with.

```{r}
mode(x)
mode(Logical)
mode(Character_Data)
is(x)
is(Logical)
is(Character_Data)
```

### Vectors

A vector is a basic data structure in R. It is a set of scalars of the same data type.

We can create vectors in different ways.

One of the main ways is to use the function c() to concatenate multiple scalars together.

```{r}
x <- c(1, 5, 4, 9, 0)
x
```

We can use function typeof() to determine the data type of a vector, and we can check the length of the vector using the funtion length() .

```{r}
x <- c(1, 5, FALSE, 9, "help")
x 
typeof(x)
```

Instead of reassigning the elements of x using the function c(), we could reassign specific elements based on the index number.

```{r}
x
x[3] <- 4
x[5] <- 0
x
typeof(x)
```

Other ways to creat vectors is to use other operators and functions such as ":" operator, seq() function, and rep() function. 

```{r}
# Create vector of consecutive numbers
y <- 1:10
y

# Create vector of a sequence of numbers
# Defining number of points in an interval or step size
seq(1, 10, by = 1)
seq(1, 10, length.out = 10)

# A set of 5 numbers with value set as 3
rep(3, 5)
```

### Factors
A factor is a special type of character vector. Factors are qualitative or categorical variables that are often used in statistical modeling. To create a factor data structure, we will first create a character vector and convert it to a factor using the factor() function.

```{r}
temperature <- c("High","Medium","Low")
temperature <- factor(temperature)
```

Converting temperature character vector to a factor type creates “levels” based on the factor values (these are the values of categorical variables).

```{r}
temperature
```

### Matrices and Arrays

So far we have discussed one-dimensional objects. We can create objects of multidimensional data. Matrices are data structures that contain data values in two dimensions. An array is a matrix with more than two dimensions. Matrices and arrays are used perform efficient calculations in a computationally fast and efficient manner.

To create a matrix, we can use the matrix() function, which takes as arguments a
data vector and parameters for the number of rows and columns.

We can determine the dimensions of a matrix using the dim() function.

```{r}
mat <- matrix(c(2,6,3,8),nrow=2,ncol=2)
mat
dim(mat)
```

We can also choose to add row names and column names to the matrix.
```{r}
rownames(mat) <- c("a", "b")
colnames(mat) <- c("c", "d")
```

```{r}
mat<-matrix(c(2,6,3,8), nrow=2, ncol=2, dimnames=list(c("a","b"), c("c","d")))
```

```{r}
# Index-based subsetting
mat[1:2,1:2]

# Column/row-based subsetting
mat["a",1:2]
```

We can also create a matrix by concatenating vectors together using rbind() function to concatenate by rows or cbind() function to concatenate by columns.

```{r}
x <- 1:3
y <- 4:6
# Combine by rows
a <- rbind(x,y)
a
# Combined by columns
b <- cbind(x,y)
b
```

To create an array, we can use the function array(), which takes as arguments vectors as input and uses the values in the dim parameter to create an array. 
```{r}
vector1 <- c(1,2,3)
vector2 <- c(5,6,7,8,9,10)

# Create an array with dimension (3,3,2) that creates 2 arrays each with 3 rows and 3 columns. 
array1 <- array(c(vector1,vector2),dim = c(3,3,2))
array1
```

### Lists
Lists are data objects which contain elements of different types including numbers, strings, vectors, and other lists. A list can also contain a matrix or even a function as its elements. 

```{r}
#Create a list of different data types

list_data <- list(c(2,4,6,8), "Hello", matrix(c(11,12,13,14), nrow=2,ncol=2), TRUE, 62.13, FALSE)
print(list_data)

# Give names to the elements in the list
names(list_data) <- c("Vector1", "Character1", "Matrix1", "Logical1", "Numeric", "Logical2")
list_data
```

We can use the function str() to list the underlying structure of the data object.
```{r}
str(list_data)
```

### Data Frames

A data frame is a table in which each column contains values of one variable or vector and each row contains one set of values from each column. Within each column, all data elements must be of the same data type. However, different columns can be of different data types. The data stored in a data frame can be of numeric, factor or character type. In addition, each column should contain same number of data elements.

To create a data frame, we can use the function data.frame():

```{r}
#Create a data frame with employee ID, salaries, and start dates

emp.data <- data.frame( 
emp_id = c("U974","U503","U298","U545","U612"),
salary = c(623.3,515.2,611.0,729.0,843.25), 
start_date = as.Date(c("2012-01-01", "2013-09-23", "2014-11-15", "2014-05-11",
  "2015-03-27")),
stringsAsFactors = FALSE
)

emp.data
```

We can use the function str() to list the underlying structure of the data object.
```{r}
str(emp.data)
```

We can extract data from the data frame and also add data to the data frame.

```{r}
#Extract salary information
emp.data$salary

#Add column vector
emp.data$dept <- c("IT","Operations","IT","HR","Finance")
```

### Conditional Statements
Logical and relational operators can be used to execute code based on certain conditions. Common operators include:
```{r}
q <- 3
t <- 5

#if else conditional statement 
if (q<t) {
    w<-q+t
} else {
    w<-q-t
}

w
```

```{r}
a<-2
b<-3
c<-4

# Using and to test two conditions, both true
if (a<b & b<c) x<-a+b+c

x
```

### Loops

We can use looping to efficiently repeat code without having to write the same code over and over.

The while loop repeats a condition while the expression in parenthesis holds true and takes the form:

```{r}
x<-0
while (x<=5){
    x <- x+1
}
x
```

For loops are used to iterate through a process a specified number of times. A
counter variable such as "i" is used to count the number of times the loop is executed: 

```{r}
#Create empty vector
y <- vector(mode="numeric")

#Loop through 1 to 10 to add values to y
for (i in 1:10) {
  y[i]<-i
}

y
```

## Advanced
### Tidyverse

One of the best advantages of R is it's unique ability to work with data. [Tidyverse](https://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf) is a collection of packages for wrangling and visualizing data for statistical analysis. These are the packages you will likely use on a day to day basis in bioinformatics.

```{r}
library(magrittr)
library(dplyr)
library(tidyr)
library(purrr)
library(janitor)
library(glue)
library(reshape2)
library(ggplot2)
```

Or just call them all...

```{r}
library(tidyverse)
```

```{r include=FALSE}
library(palmerpenguins)
df <- penguins
# What if the names were ugly?
colnames(df) <- c("Species", "ISLAND", "Bill_Length$(MM)", "bill.depth.mm", "flipper.length.mm", "body.mass.(g)", "SEX", "YEAR")
head(df)
dim(df)
```

### `janitor`

Look at how ugly these column names are...

```{r}
head(df)
```

Clean them with `janitor`

```{r}
df <- janitor::clean_names(df)
colnames(df)
head(df)
```

Get a summary of the data

```{r}
summary(penguins)
```

### `magrittr` 

**`%>%`**

Here we want to find the average bill length in inches.

```{r}
# Fine
mean_bill_length_in <- mean(df$bill_length_mm * 0.0393701, na.rm=TRUE)
print(mean_bill_length_in)
# Better
mean_bill_length_in <- df %>%
  dplyr::pull(bill_length_mm) %>%
  magrittr::multiply_by(0.0393701) %>%
  mean(na.rm=TRUE)
print(mean_bill_length_in)
```

**`set_colnames()`**
```{r}
df %>%
magrittr::set_colnames(paste0("Column_", seq(ncol(.)))) %>%
head()
```

You can also insert short hand functions!

```{r}
df %>%
dplyr::pull(bill_length_mm) %>%
(function(x) {
  return(x * 0.0393701)
}) %>%
mean(na.rm=TRUE)
```

### `dplyr`

Please check out the [full reference](https://dplyr.tidyverse.org/reference/index.html) for more functions

**`mutate()`**

Modify a new or existing column

```{r}
df %>%
dplyr::mutate(bill_length_in = bill_length_mm * 0.0393701) %>%
head()
```

**`filter()`**

Filter rows

```{r}
df %>%
dplyr::filter(species == "Chinstrap") %>%
head()
```

**`select()`**

Select columns

```{r}
df %>%
dplyr::select(species, body_mass_g) %>%
head()
```

**`pull()`**

Select columns and flatten to vector

```{r}
df %>%
dplyr::pull(body_mass_g) %>%
head()
```

**`arrange()`**

Order rows

```{r}
df %>%
dplyr::arrange(desc(body_mass_g)) %>%
head()
```

**`across()`**

Apply function across columns

```{r}
df %>%
dplyr::mutate(dplyr::across(c(bill_length_mm, bill_depth_mm), round)) %>%
head()
```

**`group_by()` + `summarise()`**
Group by columns and compute summary statistics

```{r}
df %>%
dplyr::group_by(species) %>%
dplyr::summarise(n=n(), mass=mean(body_mass_g, na.rm=TRUE))
```

**Conditional mutate**
```{r}
df %>%
dplyr::mutate(body_size=
      dplyr::case_when(body_mass_g > 5000 ~ "Massive",
                       body_mass_g > 4000 ~ "Big",
                       TRUE ~ "Small")) %>%
dplyr::pull(body_size) %>%
table()
```

### `purrr`

**Mutate with two or more columns**
```{r}
df.xyz <- data.frame(x=c(1,3), y=c(1,3), z=c(1,3))
head(df.xyz)
df.xyz <- df.xyz %>%
      dplyr::mutate(xyz = purrr::pmap(list(x, y, z), function(a, b, c) {
          a + b + c
      }))
head(df.xyz)
```

### `glue`

String manipulation and formatting

```{r}
name <- "Fred"
glue('My name is {name}.')
```

```{r}
df %>%
dplyr::select(species, bill_length_mm, sex) %>%
tidyr::drop_na() %>%
dplyr::mutate(desc = glue("This {sex} penguin ({species}) has a bill length of {bill_length_mm} mm")) %>%
head()
```

### `reshape2`

**`melt()`**

Melt a data frame for visualization

```{r}
df %>%
dplyr::select(species, bill_length_mm, bill_depth_mm) %>%
reshape2::melt(id=c("species")) %>%
head(5)
```

### `ggplot2`

Melted data frames are good for plotting

```{r warning=FALSE}
df %>%
dplyr::select(species, bill_length_mm, bill_depth_mm) %>%
reshape2::melt(id=c("species")) %>%
ggplot(aes(x=variable, y=value, color=species)) +
labs(x="", y="Length (mm)") +
geom_boxplot()
```
